---
title: "Code Challenge"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, messages = FALSE, warning = FALSE)
```



```{r message = FALSE}
library(tidyverse)
```

# Task 1

```{r}
datanames <- list.files(pattern ="*ERCOT*", full.names = TRUE)
csvdata = lapply(datanames, read_csv)

histdata2016 <- data.frame(csvdata[1])
histdata2017 <- data.frame(csvdata[2])
histdata2018 <- data.frame(csvdata[3])
histdata2019 <- data.frame(csvdata[4])
```


```{r}
histdataframe <- rbind(histdata2016, histdata2017, histdata2018, histdata2019)
```


```{r}
histdata <- histdataframe %>%
  mutate(Date = lubridate::ymd_hms(Date),
         Year = lubridate::year(Date),
         Month = lubridate::month(Date),
         Day = lubridate::day(Date),
         Hour = lubridate::hour(Date))
head(histdata)
```


# Task 2


```{r}
grouped_avg <- histdata %>%
  group_by(Year, Month, SettlementPoint) %>%
  summarise(AveragePrice = mean(Price, na.rm = T))

write.csv(grouped_avg, "AveragePriceByMonth.csv")

```
#Task 3 


This task may have some computational issues with the hourly returns, but, I did my best to continue through the problem.


```{r}
hourly_volatility <- histdata %>%
  select(Year, Hour, Price, SettlementPoint, Day, Month) %>%
  filter(Price > 0, str_detect(SettlementPoint, "HB_")) %>%
  group_by(Year, SettlementPoint, Day, Month) %>%
  #This is where I think I went wrong, HB_PAN only had a single hour for     some days, so I ahd to exclude it.
  filter(n() > 1) %>%
  mutate(hourlyreturns = log(Price) - log(lag(Price))) %>%
  drop_na() %>%
  group_by(Year, SettlementPoint) %>%
  summarise(HourlyVolatility = sd(hourlyreturns, na.rm = T))
    
write.csv(hourly_volatility, "HourlyVolatilityByYear.csv")
```
Task 3


```{r}
maxvolatility <- hourly_volatility %>%
  group_by(Year) %>%
  mutate(maxvol = max(HourlyVolatility)) %>%
  filter(HourlyVolatility == maxvol) %>%
  select(Year, SettlementPoint, HourlyVolatility)

write.csv(maxvolatility, "MaxVolatilityByYear.csv")
```


Task 4

```{r}
#Function to format data for a given settlement point
write_formatted_csv <- function(Settlement, Data) {
  

  formatteddata <- Data %>%
    filter(SettlementPoint == Settlement) %>%
    mutate(Date = lubridate::date(Date),
         Hour = Hour + 1) %>%
    select(Price, SettlementPoint, Date, Hour)%>%
    group_by(Date, SettlementPoint) %>%
    pivot_wider(names_from = Hour,
              values_from = Price)
  new_names = c("SettlementPoint", 
              "Date", 
              paste("X", colnames(formatteddata[3:24]), sep =""))

  colnames(formatteddata) = new_names
  filename = paste("formattedSpotHistory/spot_", Settlement, ".csv")
  write.csv(formatteddata, filename)
}

#Iterate through all settlements
all_settlements <- unique(histdata$SettlementPoint)
lapply(all_settlements, write_formatted_csv, histdata)
```
Bonus Task 1

```{r}
grouped_avg %>%
  mutate(ym = paste(Year, Month, sep = "-"),
         Date = lubridate::ym(ym),
         Date = format(as.Date(Date)), "%Y-%m") %>%
  filter(str_detect(SettlementPoint, "HB_")) %>%
  ggplot(aes(x = as.Date(Date), 
             y = AveragePrice, 
             color = SettlementPoint)) +
  geom_line() + 
  labs(x = "Date",
       y = "Monthly Average Price",
       title = "Monthly Average Electricity Prices Over Time",
       subtitle = "For Hubs")
```


```{r}
grouped_avg %>%
  mutate(ym = paste(Year, Month, sep = "-"),
         Date = lubridate::ym(ym),
         Date = format(as.Date(Date)), "%Y-%m") %>%
  filter(!str_detect(SettlementPoint, "HB_")) %>%
  ggplot(aes(x = as.Date(Date), 
             y = AveragePrice, 
             color = SettlementPoint)) +
  geom_line() + 
  labs(x = "Date",
       y = "Monthly Average Price",
       title = "Monthly Average Electricity Prices Over Time",
       subtitle = "For Load Zones")
```

Bonus Task 2


```{r}
hourly_volatility %>%
  ggplot(aes(x = Year,
             y = HourlyVolatility,
             color = SettlementPoint)) +
  geom_line() +
  labs(x = "Year",
       y = "Hourly Volatility",
       title = "Hourly Volatility By year",
       subtitle = "For Hubs")

```
Bonus Task 3

```{r}
#Function to find the shape profile for a given settlement point
shapeprofile <- function(Settlement, Data) {
  shapedata <- Data %>%
    filter(SettlementPoint == Settlement) %>%
    group_by(Month, Day) %>%
    mutate(avgprice = mean(Price),
           standprice = Price - (avgprice - 1)) %>%
    group_by(Day, Month, Hour) %>%
    summarise(ShapeProfValue = mean(standprice))
  filename = paste("hourlyShapeProfiles/profile_", Settlement, ".csv")
  write.csv(shapedata, filename)
}

#Iterate through all settlements
lapply(all_settlements, shapeprofile, histdata)
```


I did not have an opportunity to get to the open-ended analysis. I was planning on making an animated graph with Shiny that would convey similar information to the graphs in Bonus Task 1 in a more creative and interactive way.



